Red Wine Quality Analysis by Payam Gerami
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2) 
library(gridExtra) 
library(knitr)
library(dplyr)
library(RColorBrewer)
library(reshape2)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```

This data set contains 1,599 red wines with 11 variables on the chemical
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

Our dataset consists of threen variables with 1599 observations. here is the 
list of all the variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# exploring the variables
str(wine)
names(wine)
```

In this analysis we want to find out how each variable affects the one variable
that we most care about which is quality. Let's have a look at the values we 
have in quality variable and three other variables (alcohol, residual.sugar and 
fixed.acidity) to get a better understanding of the data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# exploring the values of four variables
summary(wine)
```

###Quality

We can see in the following diagrams that the quality distribution is fairly 
normal and we do not have sgnificant outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the quality distribution
p1 <- ggplot( aes( x = quality), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = quality), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- ggplot( aes( x = 1, y = quality), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

In the following section we explore other variables that have an effect on the 
quality.

###Volatile Acidity
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the volatile.acidity distribution
p1 <- ggplot( aes( x = volatile.acidity), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = volatile.acidity), data = wine) +
  geom_histogram() +
  xlim(0.15,0.95)
p3 <- ggplot( aes( x = 1, y = volatile.acidity), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

The amount of acetic acid in wine, which at too high of levels can lead to an 
unpleasant, vinegar taste. As we can see there are some outliers that we can 
check later to see the quality of the wine with such a high level of acetic 
acid.

###Citric Acid
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the citric.acid distribution
p1 <- ggplot( aes( x = citric.acid), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = citric.acid), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- ggplot( aes( x = 1, y = citric.acid), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

Found in small quantities, citric acid can add 'freshness' and flavor to wines. 

###Residual Sugar
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the residual.sugar distribution
p1 <- ggplot( aes( x = residual.sugar), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = residual.sugar), data = wine) +
  geom_histogram() +
  xlim(1,4)
p3 <- ggplot( aes( x = 1, y = residual.sugar), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

The amount of sugar remaining after fermentation stops, it's rare to find wines 
with less than 1 gram/liter and wines with greater than 45 grams/liter are 
considered sweet. The maximum amount of sugar we have in our dataset is 15.5 
grams but majarity of wines in our sample have between 1 and 4 grams of sugar 
and the distribution is right skewed.

###Total Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the total.sulfur.dioxide distribution
p1 <- ggplot( aes( x = total.sulfur.dioxide), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth=50)
p3 <- ggplot( aes( x = 1, y = total.sulfur.dioxide), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes 
evident in the nose and taste of wine. As we can see here we have a couple of 
outliers as well which we can check later to find out the possible effect they 
have a the quality.

###PH
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the pH distribution
p1 <- ggplot( aes( x = pH), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = pH), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- ggplot( aes( x = 1, y = pH), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 
(very basic); most wines are between 3-4 on the pH scale. The distribution is 
almost normal and most of the values are in the expected range.

###Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the alcohol distribution
p1 <- ggplot( aes( x = alcohol), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = alcohol), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- ggplot( aes( x = 1, y = alcohol), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

The percent alcohol content of the wine. Here we might have only one type of 
wine with too much alcohol in it.

# Univariate Analysis

### What is the structure of your dataset?

There are 1,599 wines in the dataset with 11 features (fixed.acidity, 
volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulphates, alcohol ).

Other observations:  

  * The quality is an integer number with 6 values.
  * Sugar content distribution is interesting and it shows that most of the 
  wines have between 1 and 4 gram/liter but there are several other wines that 
  have more than 4 grams which is not necessarly a bad thing.
  * As stated in the dataset description, it seems that diffrent types of acid 
  have significant impact on the quality.
  * 75% of the wines have less than 11.10% alcohol 

### What is/are the main feature(s) of interest in your dataset?

Based on the dataset description, it appears that the volatile acidity variable
has the most impact on the quality

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

residual sugar, alcohol and total sulfur dioxide likely contribute to the 
quality of a wine.

### Did you create any new variables from existing variables in the dataset?

Since the quality has only 6 values, we categorize this variable into the 
following factor:  

  * < 3 very bad
  * 4   bad
  * 5   average
  * 6   good
  * 7   excelent
  * \> 8 very excelent
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
wine$quality.bucket <- cut(wine$quality, 
                       breaks = c(0, 3, 4, 5, 6, 7, 8), 
                       labels = c('very bad','bad','average','good','excelent',
                                  'very excelent')) 
```


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The suger distribution is right skewed with many outliers.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=11, fig.width=9}
ggpairs(data = wine, columns=2:13)
```

Different types of acids in the wine tend to correlate with each other. Also, 
quality seems to have a semi strong correlation with alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2)
```

about volatile.acidity and alcohol relation, most important factors

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = volatile.acidity, y = residual.sugar, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2)
```

about volatile.acidity and sugar, sugar does not have a role

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = alcohol, y = residual.sugar, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2)
```

about alcohol and sugar, sugar does not have a role

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
ggplot(aes(x = quality, y = volatile.acidity, color = quality.bucket), 
       data = wine) +
  geom_boxplot()
```

less acid better

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
ggplot(aes(x = quality, y = alcohol, color = quality.bucket), 
       data = wine) +
  geom_boxplot()
```

more alcohol better

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
ggplot(aes(x = quality, y = residual.sugar, color = quality.bucket), 
       data = wine) +
  geom_boxplot()
```

sugar does not have an impact

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!