Red Wine Quality Analysis by Payam Gerami
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2) 
library(gridExtra) 
library(knitr)
library(dplyr)
library(RColorBrewer)
library(reshape2)
library(GGally)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```

This data set contains 1,599 red wines with 11 variables on the chemical
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

Our dataset consists of threen variables with 1599 observations. here is the 
list of all the variables.

```{r echo=FALSE, Univariate_Plots}
# exploring the variables
str(wine)
names(wine)
```

In this analysis we want to find out how each variable affects the one variable
that we most care about which is quality. Let's have a look at the values we 
have in quality variable and three other variables (alcohol, residual.sugar and 
fixed.acidity) to get a better understanding of the data.

```{r, echo=FALSE}
# exploring the values of four variables
summary(wine$quality)
summary(wine$alcohol)
summary(wine$residual.sugar)
summary(wine$fixed.acidity)
```

We can see in the following diagrams that the quality distribution is fairly 
normal and we do not have isgnificant outliers.

```{r, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the quality distribution
p1 <- ggplot( aes( x = quality), data = wine) +
  geom_histogram()
p2 <- ggplot( aes( x = quality), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- ggplot( aes( x = 1, y = quality), data = wine) +
  geom_boxplot()
grid.arrange(p1,p2,p3,ncol = 3)
```

Other distributions

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!