Red Wine Quality Analysis by Payam Gerami
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2) 
library(gridExtra) 
library(knitr)
library(dplyr)
library(RColorBrewer)
library(reshape2)
library(GGally)
library(corrplot)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```

This data set contains 1,599 red wines with 11 variables on the chemical
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent).

# Univariate Plots Section

Our dataset consists of threen variables with 1599 observations. here is the 
list of all the variables.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# exploring the variables
str(wine)
names(wine)
```

In this analysis we want to find out how each variable affects the one variable
that we most care about which is quality. Let's have a look at the values we 
have in quality variable and three other variables (alcohol, residual.sugar and 
fixed.acidity) to get a better understanding of the data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# exploring the values of four variables
summary(wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
create_histogram <- function(variable, binwidth = 0.5) {
  return(ggplot(aes_string(x = variable), data = wine) + 
           geom_histogram(binwidth = binwidth))}
create_boxplot <- function(variable) {
  return(ggplot(aes_string(x = variable, y = variable), data = wine) + 
           geom_boxplot())}
```

###Quality

We can see in the following diagrams that the quality distribution is fairly normal and we do not have sgnificant outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
# exploring the quality distribution
p1 <- create_histogram('quality')
p2 <- create_boxplot('quality')
grid.arrange(p1,p2,ncol = 2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$quality)
```

In the following section we explore other variables that have an effect on the 
quality.

### Fixed Acidity
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the fixed.acidity distribution
p1 <- create_histogram('fixed.acidity')
p2 <- ggplot( aes( x = fixed.acidity), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('fixed.acidity')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$fixed.acidity)
```

Most acids involved with wine or fixed or nonvolatile - do not evaporate readily
 (tartaric acid - g / dm^3).

### Volatile Acidity
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the volatile.acidity distribution
p1 <- create_histogram('volatile.acidity', 0.01)
p2 <- ggplot( aes( x = volatile.acidity), data = wine) +
  geom_histogram() +
  xlim(0.15,0.95)
p3 <- create_boxplot('volatile.acidity')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$volatile.acidity)
```

The amount of acetic acid in wine, which at too high of levels can lead to an 
unpleasant, vinegar taste. As we can see there are some outliers that we can 
check later to see the quality of the wine with such a high level of acetic 
acid (acetic acid - g / dm^3).

### Citric Acid
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the citric.acid distribution
p1 <- create_histogram('citric.acid', 0.02)
p2 <- ggplot( aes( x = citric.acid), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('citric.acid')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$citric.acid)
```

Found in small quantities, citric acid can add 'freshness' and flavor to wines 
(g / dm^3). 

### Residual Sugar
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the residual.sugar distribution
p1 <- create_histogram('residual.sugar')
p2 <- ggplot( aes( x = residual.sugar), data = wine) +
  geom_histogram() +
  xlim(1,4)
p3 <- create_boxplot('residual.sugar')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$residual.sugar)
```

The amount of sugar remaining after fermentation stops, it's rare to find wines 
with less than 1 gram/liter and wines with greater than 45 grams/liter are 
considered sweet. The maximum amount of sugar we have in our dataset is 15.5 
grams but majarity of wines in our sample have between 1 and 4 grams of sugar 
and the distribution is right skewed (g / dm^3).

### Chlorides
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the chlorides distribution
p1 <- create_histogram('chlorides',0.01)
p2 <- ggplot( aes( x = chlorides), data = wine) +
  geom_histogram() +
  xlim(0.05,0.2)
p3 <- create_boxplot('chlorides')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$chlorides)
```

The amount of salt in the wine (sodium chloride - g / dm^3).

### Free Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the free.sulfur.dioxide distribution
p1 <- create_histogram('free.sulfur.dioxide', 1)
p2 <- ggplot( aes( x = free.sulfur.dioxide), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('free.sulfur.dioxide')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$free.sulfur.dioxide)
```

The free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved
gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine 
(mg / dm^3).

### Total Sulfur Dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the total.sulfur.dioxide distribution
p1 <- create_histogram('total.sulfur.dioxide', 2)
p2 <- ggplot( aes( x = total.sulfur.dioxide), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('total.sulfur.dioxide')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$total.sulfur.dioxide)
```

Amount of free and bound forms of S02; in low concentrations, SO2 is mostly 
undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes 
evident in the nose and taste of wine. As we can see here we have a couple of 
outliers as well which we can check later to find out the possible effect they 
have a the quality (mg / dm^3).

### Density
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the density distribution
p1 <- create_histogram('density', 0.001)
p2 <- ggplot( aes( x = density), data = wine) +
  geom_histogram() +
  xlim(0.9956, 0.9978)
p3 <- create_boxplot('density')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$density)
```

the density of water is close to that of water depending on the percent alcohol 
and sugar content (g / cm^3).

###PH
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the pH distribution
p1 <- create_histogram('pH', 0.02)
p2 <- ggplot( aes( x = pH), data = wine) +
  geom_histogram(binwidth = 0.005) +
  xlim(3.210,3.400)
p3 <- create_boxplot('pH')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$pH)
```

Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 
(very basic); most wines are between 3-4 on the pH scale. The distribution is 
almost normal and most of the values are in the expected range.

### Sulphates 
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the sulphates distribution
p1 <- create_histogram('sulphates', 0.01)
p2 <- ggplot( aes( x = sulphates), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('sulphates')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$sulphates)
```

A wine additive which can contribute to sulfur dioxide gas (S02) levels, wich 
acts as an antimicrobial and antioxidant (potassium sulphate - g / dm3).

### Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}
# exploring the alcohol distribution
p1 <- create_histogram('alcohol', 0.1)
p2 <- ggplot( aes( x = alcohol), data = wine) +
  geom_histogram() +
  scale_x_log10()
p3 <- create_boxplot('alcohol')
grid.arrange(p1,p2,p3,ncol = 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine$alcohol)
```

The percent alcohol content of the wine. Here we might have only one type of 
wine with too much alcohol in it (% by volume).

# Univariate Analysis

### What is the structure of your dataset?

There are 1,599 wines in the dataset with 11 features (fixed.acidity, 
volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulphates, alcohol ).

Other observations:  

  * The quality is an integer number with 6 values.
  * Sugar content distribution is interesting and it shows that most of the 
  wines have between 1 and 4 gram/liter but there are several other wines that 
  have more than 4 grams which is not necessarly a bad thing.
  * As stated in the dataset description, it seems that diffrent types of acid 
  have significant impact on the quality.
  * 75% of the wines have less than 11.10% alcohol 

### What is/are the main feature(s) of interest in your dataset?

Based on the dataset description, it appears that the volatile acidity variable
has the most impact on the quality

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

residual sugar, alcohol and total sulfur dioxide likely contribute to the 
quality of a wine.

### Did you create any new variables from existing variables in the dataset?

Since the quality has only 6 values, we categorize this variable into the 
following factor:  

  * < 3 very bad
  * 4   bad
  * 5   average
  * 6   good
  * 7   excelent
  * \> 8 very excelent
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
wine$quality.bucket <- cut(wine$quality, 
                       breaks = c(0, 3, 4, 5, 6, 7, 8), 
                       labels = c('very bad','bad','average','good','excelent',
                                  'very excelent')) 
```


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The suger distribution is right skewed with many outliers.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=9}
M <- cor(wine[, 2:13])
corrplot(M, method="number")
```

Different types of acids in the wine tend to correlate with each other. Also, 
quality seems to have a semi strong correlation with alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = quality.bucket, y = volatile.acidity), data = wine) + 
  geom_jitter(alpha = 0.5, size = 2)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

The general trend between the volatile.acidity and the quality provides an 
evidence that as volatile acidity decreases, the quality increases.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = quality.bucket, y = alcohol), data = wine) + 
  geom_jitter(alpha = 0.5, size = 2)  +
  geom_boxplot( alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

And the general trend between the alcohol and the quality provides an evidence 
that as alcohol increases, the quality increases as well.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = quality.bucket, y = residual.sugar), data = wine) + 
  geom_jitter(alpha = 0.5, size = 2)  +
  geom_boxplot( alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

And the general trend between residual.sugar and the quality confirms that 
residual sugar has a minimal impact on the quality of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
ggplot(aes(x = fixed.acidity, y = pH), data = wine) +
  geom_jitter(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE,size=1)
```

The strongest correlation is between fixed.acidity and pH. As it's clear in the 
diagram, fixed.acidity decreases when pH decreases.

# Bivariate Analysis

### Some of the Observed Relationships

We can infer that higher quality wines tend to have high alcohol percentage and 
low volatile acidity.

### Interesting relationships between the other features

It appears that sugar does not have a clear impact on quality. 

### What was the strongest relationship you found?

The relationship between volatile acidity and quality is very strong. The plot 
clearly shows as the quality goes up, the volatile acidity level decreases.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))
```

Volatile.acidity and Alcohol seems to be the most important factors affecting 
the quality of the wine. It looks like wines with less amount of volatile 
acidity and more amount of alcohol have better quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}
ggplot(aes(x = sulphates, y = citric.acid, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2)+
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))
```

Two other features that have impact on the quality but not as much as alcohol 
and volatile acidity are sulphates and citric acid. As we can see, when the 
combination of citric acid and sulphates increases, we tend to have a better 
quality wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}
ggplot(aes(x = fixed.acidity, y = pH, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2)+
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))
```

Even though pH and fixed.acidity are strongly correlated, there does not seem 
to be any pattern between the combination of the two and the quality.

# Multivariate Analysis

### Relationships observed in investigation

The main features of interest are acidity and alcohol.

### Interesting or surprising interactions between features?

between other variables and quality, it seems that no clear correlation is 
available.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.bucket),
       data = wine) + 
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'color', 
                            override.aes = list(alpha = 1, size = 2))) +
  geom_jitter(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  xlab('volatile acidity (acetic acid - g / dm^3)') +
  ylab('alcohol (% by volume)') +
  labs(title = "Volatile Acidity vs. Alcohol by Quality Categories") + 
  theme(plot.title = element_text(hjust = 0.5))
```

### Description One

This diagram illustrates the correlation between the two most effective features
on the quality. Wine seems to have better quality when volatile acidity is 
between 0.3 and 0.4 g / dm^3, and alcohol is between 11 and 14 percent.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=9}
ggplot(aes(x = quality.bucket, y = volatile.acidity), 
       data = wine) +
  geom_jitter(alpha = 0.5, size = 2)  +
  geom_boxplot( alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab('Wine Quality') +
  ylab('volatile acidity (acetic acid - g / dm^3)') +
  labs(title = "Volatile Acidity by Quality Categories") + 
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Two

Volatile acidity has the strongest correlation with quality. The above boxplot 
clearly shows the increasing trend in the volatile acidity when wine quality 
moves from lower level to higher level.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=9}
ggplot(aes(x = quality.bucket, y = alcohol), 
       data = wine) +
  geom_jitter(alpha = 0.5, size = 2)  +
  geom_boxplot( alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab('Wine Quality') +
  ylab('alcohol (% by volume)') +
  labs(title = "Alcohol by Quality Categories") + 
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Three

Alcohol percentage has the second most strongest correlation with the quality of
the wine. It is clear in the above boxplot that high quality wines tend to have 
higher alcohol percentage.

------

# Reflection

In this analysis I explored the effect of different features of red wine on it's
quality based on the data set that contained 1,599 red wines with 11 variables 
on the chemical properties. I categorized the quality feature into 6 categories 
and by exploring the correlation between the features I realized that the most 
effective features are volatile acidity and alcohol percentage. By looking at 
combination of other features it's really hard to find a repeatable pattern and 
as a result I was not able to find the perfect balance between the features that
would result in perfect quality wine.  

To expand this analysis, more features could be combined together. There might 
exists a perfect balance between 5 features together that would result in a 
perfect quality wine. It is also important to keep in mind that we could 
categorize the red wines into different types based on characteristics sush as 
Sweetness, Acidity, Tannin, Fruit and Body then analyse the data to find the 
perfect balance for each type. This approach most probably would result in 
better understanding of the features.

  